The Unicode Cookbook \linebreak for Linguists

### 2018-06-08 book/chapters/introduction.tex
deals with paperhivers through chapter 2
deals with all paperhiver comments

When considering worldwide linguistic diversity, including the many
lesser-studied and endangered languages, there exist numerous different
orthographies using symbols from the same scripts. For example, there are
hundreds of orthographies using Latin-based alphabetic scripts. All of these
orthographies use the same symbols, but these symbols differ in meaning and
usage throughout the various orthographies. To be able to computationally use
and compare different orthographies, we need a way to specify all orthographic
idiosyncrasies in a computer-readable format. We call such specifications
\textsc{orthography profiles}. Ideally, these specifications have to be
integrated into so-called Unicode Locales,\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
though we will argue that in practice this is often not the most useful solution for the kind of problems
arising in the daily practice of many linguists. Consequently, a central goal of
this book is to flesh out the linguistic challenges for locale descriptions,
and work out suggestions to improve their structure for usage in a linguistic
context. Conversely, we also aim to improve linguists' understanding and
appreciation for the accomplishments of the Unicode Consortium in the
development of the Unicode Standard.

The need to use computational methods to compare different orthographies arises most
forcefully in the context of language comparison. Concretely, the proper
processing of orthographies and transcription systems becomes crucial for the
development of quantitative methods for language comparison and historical
reconstruction, in order to investigate worldwide linguistic variation and to
model the historical and areal processes that underlie linguistic
diversity. In this work, it is crucial to be able to flexibly process numerous
resources in different orthographies. In many cases even different resources
on the same language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus, a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

@@

Breaking it down further, a script consists of \textsc{graphemes}, which are writing 
system-specific minimally distinctive symbols (see below). Graphemes may consist of one or more 
\textsc{characters}. The term \textsc{character} is overladen. In the linguistic terminology of writing
systems, a \textsc{character} is a general term for any self-contained element
in a writing system. A second interpretation is used as a conventional term for a unit in the Chinese writing
system \citep{Daniels1996}. In technological terminology, a \textsc{character} 
refers to the electronic encoding of a component in a writing system that has semantic 
value (see Section \ref{character-encoding-system}). Thus in this work we must navigate 
between the general linguistic and technological terms for \textsc{character} 
and \textsc{grapheme} because of how these notions are defined and how they relate at the intersection 
between the International Phonetic Alphabet and the Unicode Standard (Chapter \ref{ipa-meets-unicode}).

@@

For that reason, it is better to analyze an orthography as a collection of
graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
particular writing system. It was modeled
after the term \textsc{phoneme} (an abstract representation of
a distinct sound in a specific language) and as such it represents a contrastive graphical unit in a
writing system (see \citealp{Kohrt1986} for a historical overview of the term
\textsc{grapheme}). Most importantly, a single grapheme regularly consists of multiple
characters, like <th>, <ou> and <gh> in English (note that each character in
these graphemes is also a separate grapheme in English). Such complex graphemes
are often used to represent single phonemes. So, a combination of characters is
used to represent a single phoneme. Note that the opposite is also found in
writing systems, in cases in which a single character represents a combination
of two or more phonemes. For example, <x> in English orthography sometimes represents a
combination of the phonemes /k/ and /s/, as in the word `index' [ˈɪnˌdɛks]; other times 
it is pronounced as /z/, such as in the words `Xerox' [ˈzɪrˌɑks] and `example' [ɪɡˈzæmpəl]. 
As one can see, there can be non-trivial mappings between graphemes and phonemes in 
orthographies like English, e.g.\ `give', `gin', `jingle', where the graphemes 
<g> and <j> and the phonemes /g/ and /dʒ/ have a complex mapping.

@@

\noindent This exposition of the linguistic terminology involved in describing writing
systems has been purposely brief. We have highlighted some of the linguistic
notions that are pertinent to, yet sometimes confused with, the technological
definitions developed for the computational processing of the world's writing
systems, which we describe in the next Chapter.

>>>>

When considering worldwide linguistic diversity, including the many
lesser-studied and endangered languages, there exist numerous different
orthographies using symbols from the same scripts. For example, there are
hundreds of orthographies using Latin-based alphabetic scripts. All of these
orthographies use the same symbols, but these symbols differ in meaning and
usage throughout the various orthographies. To be able to computationally use
and compare different orthographies, we need a way to specify all orthographic
idiosyncrasies in a computer-readable format. We call such specifications
\textsc{orthography profiles}. Ideally, these specifications have to be
integrated into so-called Unicode Locales,\footnote{\url{http://cldr.unicode.org/locale_faq-html}} 
though we will argue that in practice this is often not the most useful solution for the kind of problems
arising in the daily practice of many linguists. Consequently, a central goal of this book is to flesh out the linguistic-specific challenges regarding Unicode Locales and to work out suggestions to simplify their structure for usage in a linguistic context. Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.

The need to use computational methods to compare different orthographies arises most
forcefully in the context of language comparison. Concretely, the proper
processing of orthographies and transcription systems becomes critical for the
development of quantitative methods for language comparison and historical
reconstruction. In order to investigate worldwide linguistic variation and to
model the historical and areal processes that underlie linguistic
diversity, it is crucial that we are able to flexibly process numerous
resources in different orthographies. In many cases even different resources
on the same language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus, a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

@@

Breaking it down further, a script consists of \textsc{graphemes}, which are writing 
system-specific minimally distinctive symbols (see below). Graphemes may consist of one or more 
\textsc{characters}. The term \textsc{character} is overladen. In the linguistic terminology of writing
systems, a \textsc{character} is a general term for any self-contained element
in a writing system. A second interpretation is used as a conventional term for a unit in the Chinese writing
system \citep{Daniels1996}. In technical terminology, a \textsc{character} 
refers to the electronic encoding of a component in a writing system that has semantic 
value (see Section \ref{character-encoding-system}). Thus in this work we must navigate 
between the general linguistic and technical terms for \textsc{character} 
and \textsc{grapheme} because of how these notions are defined and how they relate at the intersection 
between the International Phonetic Alphabet and the Unicode Standard (Chapter \ref{ipa-meets-unicode}).

@@

For that reason, it is better to analyze an orthography as a collection of
graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
particular writing system. It was modeled
after the term \textsc{phoneme} (an abstract representation of
a distinct sound in a specific language) and as such it represents a contrastive graphical unit in a
writing system (see \citealp{Kohrt1986} for a historical overview of the term
\textsc{grapheme}). Most importantly, a single grapheme regularly consists of multiple
characters, like <th>, <ou> and <gh> in English (note that each character in
these graphemes is also a separate grapheme in English). Such complex graphemes
are often used to represent single phonemes. So, a combination of characters is
used to represent a single phoneme. Note that the opposite is also found in
writing systems, in cases in which a single character represents a combination
of two or more phonemes. For example, <x> in English orthography sometimes represents a
combination of the phonemes /k/ and /s/, as in the word `index' [ˈɪnˌdɛks]; other times 
it is pronounced as /z/, such as in the words `Xerox' [ˈzɪrˌɑks]; and in `example' [ɪɡˈzæmpəl] 
it is a combination of /g/ and /s/. As one can see, there can be non-trivial mappings between graphemes and phonemes in 
orthographies like English, e.g.\ `give', `gin', `jingle', where the graphemes 
<g> and <j> and the phonemes /g/ and /dʒ/ have a complex mapping.

@@

\noindent This exposition of the linguistic terminology involved in describing writing
systems has been purposely brief. We have highlighted some of the linguistic
notions that are pertinent to, yet sometimes confused with, the technical
definitions developed for the computational processing of the world's writing
systems, which we describe in the next Chapter.

### 2018-06-08 book/chapters/orthography_profiles.tex
deals with all paperhiver comments

As an example of graphemic tokenization, note the three different levels of
technological and linguistic elements that interact in the hypothetical lexical
form <tsʰṍ̰shi>:

>>>>

As an example of graphemic tokenization, note the three different levels of
technical and linguistic elements that interact in the hypothetical lexical
form <tsʰṍ̰shi>:

### 2018-06-08 book/chapters/pitfalls.tex
deals with all paperhiver comments

In this chapter we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; rather we aim to highlight where the
technological aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been made differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call such behavior that
executes without error but does something different than the user
expected -- often unknowingly -- a \textsc{pitfall}.

@@

There are two obstacles regarding missing glyphs. One is practical: 
designing glyphs includes many different considerations and 
it is a time-consuming process, especially when done well. 
Traditional expectations of what specific characters should look like need
to be taken into account when designing glyphs. Those expectations are often not
well documented, and it is mostly up to the knowledge and expertise of the font
designer to try and conform to them. Most designers 
produce fonts that only include glyphs for certain parts of the Unicode
Standard.

The second obstacle is technical: the maximum number of glyphs that can be 
defined by the TrueType font standard or the OpenType specification 
(ISO/IEC 14496-22:2015) is 65,535. The current version of the Unicode Standard 
contains 137,374 characters. Thus, no single font can provide individual 
glyphs for all Unicode characters.

@@

We also suggest installing at least one \textsc{fall-back
font}, which provides glyphs that show the user some information about
the underlying encoded character. Apple Macintoshes have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character.\footnote{\url{http://unifoundry.com/unifont.html}} Finally,
SIL International provides a \textsc{SIL Unicode BMP Fallback
Font}. This font does not show a real 
glyph, but instead shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode code point of the
character to be displayed.\footnote{\url{http://scripts.sil.org/UnicodeBMPFallbackFont}}

@@

In general, finding a specific character in the Unicode Standard is often non-trivial. 
The names of the blocks can help, but they are not (and were never supposed
to be) a foolproof structure. It is neither the goal nor the aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end-users available.\footnote{The Unicode
website offers a basic interface to the code charts at
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Ergonis Software, available for both Macintosh and
Windows. There are also various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net} or
\url{http://unicode-search.net}. A further useful approach for searching characters
using shape matching is \url{http://shapecatcher.com}.}

@@

For some characters, there is more than one possible encoding in the Unicode
Standard. This means that for the computer
there exists multiple different entities, which for the user, may be visually the same. This
leads to, for example, problems with search. The computer searches for specific 
code points and by design does not return all visually similar characters.
As a solution, the Unicode Standard includes a notion of \textsc{canonical
equivalence}. Different encodings are explicitly declared as equivalent in the
Unicode Standard code tables. Further, to harmonize all encodings in a specific
piece of text, the Unicode Standard proposes a mechanism of
\textsc{normalization}.

@@

The character, represented here by glyph <Å>, is encoded in the Unicode Standard
in the first two examples by a single-character sequence; each is assigned a
different code point. In the third example, the glyph is encoded in a
multiple-character sequence that is composed of two character code points. All
three sequences are \textsc{}, i.e.~they are strings that
represent the same abstract character and because they are not distinguishable
by the user, the Unicode Standard requires them to be treated the same in
regards to their behavior and appearance. Nevertheless, they are encoded
differently. For example, if one were to search an electronic text (with
software that does not apply Unicode Standard normalization) for
\textsc{angstrom sign} (\uni{212B}), then the instances of \textsc{latin 
capital letter a with ring above} (\uni{00C5}) would not be found.

@@

The other meaning of encoding has to do with the fact that computers represent data and instructions in patterns of bits. A bit pattern is a combination of binary digits arranged in a sequence. And how these sequences are carved up into bit patterns is determined by how they are \textit{encoded}. Thus the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems because they determine how sequences of bit patterns are determined, which are then mapped to characters. The terminological issue here is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.\ the code point, which can be encoded differently by different encoding systems.

>>>>

In this chapter we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; rather we aim to highlight where the
technical aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been made differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call such behavior that
executes without error but does something different than the user
expected -- often unknowingly -- a \textsc{pitfall}.

@@

There are two obstacles regarding missing glyphs. One is practical: 
designing glyphs includes many different considerations and 
it is a time-consuming process, especially when done well. 
Traditional expectations of what specific characters should look like need
to be taken into account when designing glyphs. Those expectations are often not
well documented, and it is mostly up to the knowledge and expertise of the font
designer to try and conform to them. Furthermore, the number of characters 
supported by Unicode is vast. Therefore, most designers 
produce fonts that only include glyphs for certain parts of the Unicode
Standard.

The second obstacle is technical: the maximum number of glyphs that can be 
defined by the TrueType font standard and the OpenType specification 
(ISO/IEC 14496-22:2015) is 65,535. The current version of the Unicode Standard 
contains 137,374 characters. Thus, no single font can provide individual 
glyphs for all Unicode characters.

@@

We also suggest installing at least one \textsc{fall-back
font}, which provides glyphs that show the user some information about
the underlying encoded character. Apple computers have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character.\footnote{\url{http://unifoundry.com/unifont.html}} Finally,
SIL International provides a \textsc{SIL Unicode BMP Fallback
Font}. This font does not show a real 
glyph, but instead shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode code point of the
character to be displayed.\footnote{\url{http://scripts.sil.org/UnicodeBMPFallbackFont}}

@@

In general, finding a specific character in the Unicode Standard is often non-trivial. 
The names of the blocks can help, but they are not (and were never supposed
to be) a foolproof structure. It is neither the goal nor the aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end-users available.\footnote{The Unicode
website offers a basic interface to the code charts at
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Ergonis Software, available for both Mac and
Windows. There are also various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net} or
\url{http://unicode-search.net}. A further useful approach for searching characters
using shape matching is \url{http://shapecatcher.com}.}

@@

For some characters, there is more than one possible encoding in the Unicode
Standard. This means that for the computer
there exists multiple different entities, which for the user, may be visually the same. This
leads to, for example, problems with search. The computer searches for specific 
code points and by design does not return all visually similar characters.
As a solution, the Unicode Standard includes a notion of \textsc{canonical
equivalence}. Different encodings are explicitly declared as equivalent in the
Unicode Standard code tables. Further, to harmonize all encodings in a specific
piece of text, the Unicode Standard proposes a mechanism of
\textsc{normalization}. The process of normalization and the 
Unicode Normalization Forms are described 
in detail in the Unicode Standard Annex \#15 online.\footnote{\url{http://unicode.org/reports/tr15/}} 
Here we provide a brief summary of that material as it pertains to canonical equivalence.

@@

\noindent The character, represented here by glyph <Å>, is encoded in the Unicode Standard
in the first two examples by a single-character sequence; each is assigned a
different code point. In the third example, the glyph is encoded in a
multiple-character sequence that is composed of two character code points. All
three sequences are \textsc{}, i.e.~they are strings that
represent the same abstract character and because they are not distinguishable
by the user, the Unicode Standard requires them to be treated the same in
regards to their behavior and appearance. Nevertheless, they are encoded
differently. For example, if one were to search an electronic text (with
software that does not apply Unicode Standard normalization) for
\textsc{angstrom sign} (\uni{212B}), then the instances of \textsc{latin 
capital letter a with ring above} (\uni{00C5}) would not be found.

@@

The other meaning of encoding has to do with the fact that computers represent data and instructions in patterns of bits. A bit pattern is a combination of binary digits arranged in a sequence. And how these sequences are carved up into bit patterns is determined by how they are \textit{encoded}. Thus the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems because they determine how sequences of bit patterns are determined, which are then mapped to characters.\footnote{UTF stands for Unicode Transformation Format. It a method for translating numbers into binary data and vice versa. There are several different UTF encoding formats, e.g. UTF-8 is a variable-length encoding that uses 8-bit code units, is compatible with ASCII, and is common on the web. UTF-16 is also variable-length, uses 16-bit code units, and is used system-internally by Windows and Java. See further discussion under \textit{Code units} in Section \ref{pitfall-file-formats}. For more in-depth discussion, refer to the Unicode Frequently Asked Questions and additional sources therein: \url{http://unicode.org/faq/utf_bom.html}.} The terminological issue here is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.\ the code point, which can be encoded differently by different encoding systems.

### 2018-06-08 book/chapters/practical_recommendations.tex
deals with all paperhiver comments

\section{For programmers and potential programmers}
If you have made it this far, and you are eager to know more about the technological aspects of the Unicode Standard and how they relate to software programming, we recommend two light-hearted blog posts on the topic. The classic blog post about what programmers should know about the Unicode Standard is Joel Spolsky's \textit{The Absolute Minimum Every Software Developer Absolutely, Positively Must Know About Unicode and Character Sets (No Excuses!)}.\footnote{\url{https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/}} A more recent blogpost, with a bit more of the technical details, is by David C. Zentgraf and is titled, \textit{What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text}.\footnote{\url{http://kunststube.net/encoding/}} This post is aimed at software developers and uses the PHP language for examples.

>>>>

\section{For programmers and potential programmers}
If you have made it this far, and you are eager to know more about the technical aspects of the Unicode Standard and how they relate to software programming, we recommend two light-hearted blog posts on the topic. The classic blog post about what programmers should know about the Unicode Standard is Joel Spolsky's \textit{The Absolute Minimum Every Software Developer Absolutely, Positively Must Know About Unicode and Character Sets (No Excuses!)}.\footnote{\url{https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/}} A more recent blogpost, with a bit more of the technical details, is by David C. Zentgraf and is titled, \textit{What Every Programmer Absolutely, Positively Needs To Know About Encodings And Character Sets To Work With Text}.\footnote{\url{http://kunststube.net/encoding/}} This post is aimed at software developers and uses the PHP language for examples.
